Is malaria control profitable? Return on investment of privately-managed residential fumigations at a large sugarcane processing facility in Southern Mozambique

### 2018-04-18 inst/rmd/01_introduction.Rmd
replacement felm

## Background Literature

@@

Regardless of the academic literature on the risk of workers becoming infected with malaria, the companies running large agricultural operations in malaria-endemic regions generally consider the disease to be an important enough threat to workers' health to merit some form of investment in malaria control [@Pluess2009]. Insecticide fumigation, bednet distribution, test-and-treat clinics, and occasional educational campaigns are common forms of corporate social responsibility to combat malaria. The evidence of malaria's negative effects on both health and wealth, at both the individual [@Cole2006] and collective [@McCarthy2000] levels, as well as in both the short [@AsensoOkyere1997] [@Ajani2010] and long [@Hong2011] [@Sachs2002] terms, are amply described in the public health and economics literature [@Phillips98]. That said, very little exists in the literature examining the costs and benefits of malaria control from a private ledger perspective (ie, the point of view a business investor). Unlike a government or individual, a private firm investing in malaria control may be most interested not in its long-term macroeconomic effects, nor its short-term personal health effects, but rather on the impact on productivity (and the extent to which that producitivty's benefits accrue to the firm), absenteeism, and the opportunity costs of expenditures in malaria control. In other words, though the large magnitude of malaria control's benefits are well known, the portion of those benefits accrued by a private firm investing in malaria control is unknown. \improvement{This is a rough transition; need to add some details here.}

A World Bank study on Nigerian sugarcane cutters found that the simple availability of testing and treatment increased productivity by 10% in the weeks following the provision of services, the conclusion being that both the treatment and the test result were effective in increasing productivity, the latter simply increasing the information available which could influence personal labor allocation decisions [@Dillon2014].

@@

This study does not endeavor to expand the current body of knowledge regarding malaria's ill effects on individuals and societies; rather, it aims to provide empirical evidence pertaining to a facet of malaria economics with very little in the literature: malaria control from a private-sector investment perspective. Our study adds to the existing literature by showing the effect of specific malaria control interventions on worker absenteeism, and translating that effect into a return on investment.

@@

Maragra (figure 1, panels iii and iv) is located in the district of Manhiça (figure 1, panel ii), a semi-rural area in the south of Mozambique (figure 1, panel i). 80 kilometers north of the Mozambican capital of Maputo, the district is low-lying, consisting largely of savannah and wetlands along the Incomati River. Most of the areas 160,000 residents [@Sacoor2013] work as subsistence farmers. Migration from the area to South Africa for the purpose of employment in the profitable construction industry is common, especially among men [@Nhacolo_2006], as is migration to the area (from other parts of Mozambique) for seasonal work on the sugarcane plantations at Maragra and the slightly larger facility in Xinavane (at the district's border with Magude) [@OLaughlin2016].

Poverty is rife in southern Mozambique, and its associated illnesses take their toll on the population. The community prevalence of HIV/AIDS is as high as 40% [@Gonzlez2012]; even in a more recent study suggesting a much lower prevalence of 22%, the risk of infection is still twice that of nearby areas [@Mocumbi2017]. Recent years have seen a three-fold increase in tuberculosis [@GarcaBasteiro2017]. Malaria, which has the greatest mortality burden due to the fact that the young are especially vulnerable to its effects, is perennial, though worse during the rainy season (December - April) [@Sate2003]. Adult malaria is essentially universal, albeit much of it asymptomatic [@Mayor2007]. In regards to worker health, malaria is Maragra's primary concern, being so important as to justify the existence of both an on-site testing laboratory and clinic, as well as a malaria control department.

>>>>

@@

Regardless of the academic literature on the risk of workers becoming infected with malaria, the companies running large agricultural operations in malaria-endemic regions generally consider the disease to be an important enough threat to workers' health to merit some form of investment in malaria control [@Pluess2009]. Insecticide fumigation, bednet distribution, test-and-treat clinics, and occasional educational campaigns are common forms of corporate social responsibility to combat malaria. A World Bank study on Nigerian sugarcane cutters found that the simple availability of testing and treatment increased productivity by 10% in the weeks following the provision of services, the conclusion being that both the treatment and the test result were effective in increasing productivity, the latter simply increasing the information available which could influence personal labor allocation decisions [@Dillon2014].

@@

The evidence of malaria's negative effects on both health and wealth, at both the individual [@Cole2006] and collective [@McCarthy2000] levels, as well as in both the short [@AsensoOkyere1997] [@Ajani2010] and long [@Hong2011] [@Sachs2002] terms, are amply described in the public health and economics literature [@Phillips98]. That said, very little exists in the literature examining the costs and benefits of malaria control from a private ledger perspective (ie, the point of view a business investor). Unlike a government or individual, a private firm investing in malaria control may be most interested not in its long-term macroeconomic effects, nor its short-term personal health effects, but rather on the impact on productivity (and the extent to which that producitivty's benefits accrue to the firm), absenteeism, and the opportunity costs of expenditures in malaria control. In other words, though the large magnitude of malaria control's benefits are well known, the portion of those benefits accrued by a private firm investing in malaria control is unknown.

This study does not endeavor to expand the current body of knowledge regarding malaria's ill effects on individuals and societies; rather, it aims to provide empirical evidence pertaining to a facet of malaria economics with very little in the literature: malaria control from a private-sector investment perspective. Our study adds to the existing literature by showing the effect of specific malaria control interventions on worker absenteeism, and translating that effect into a return on investment. Unlike previous studies on the effectiveness of malaria control interventions, our's focuses solely on one firm carrying out one intervention, takes advantage of individual-level data, and analyzes results from a ledger perspective.

@@

Maragra (figure 1, panel iii) is located in the district of Manhiça (figure 1, panel ii), a semi-rural area in the south of Mozambique (figure 1, panel i). 80 kilometers north of the Mozambican capital of Maputo, the district is low-lying, consisting largely of savannah and wetlands along the Incomati River. Most of the areas 160,000 residents [@Sacoor2013] work as subsistence farmers. Migration from the area to South Africa for the purpose of employment in the profitable construction industry is common, especially among men [@Nhacolo_2006], as is migration to the area (from other parts of Mozambique) for seasonal work on the sugarcane plantations at Maragra and the slightly larger facility in Xinavane (at the district's border with Magude) [@OLaughlin2016].

Poverty is rife in southern Mozambique, and its associated illnesses take their toll on the population. The community prevalence of HIV/AIDS is as high as 40% [@Gonzlez2012]; even in a more recent study suggesting a much lower prevalence of 22%, the risk of infection is still twice that of nearby areas [@Mocumbi2017]. Recent years have seen a three-fold increase in tuberculosis [@GarcaBasteiro2017]. Malaria, which has the greatest mortality burden due to the fact that the young are especially vulnerable to its effects, is perennial, though worse during the rainy season (December - March) [@Sate2003]. Adult malaria is essentially universal, albeit much of it asymptomatic [@Mayor2007]. In regards to worker health, malaria is Maragra's primary concern, being so important as to justify the existence of both an on-site testing laboratory and clinic, as well as a malaria control department.

Maragra workers are mostly seasonal, working for the firm approximately half of the year during harvest time, and cultivating crops, working in construction, or going unemployed (or working elsewhere) during the other half. Though many workers live "on-site" (ie, within the delineated property of the firm itself), a sizeable minority resides in the environs (figure 1, panel iv). Maragra provides indoor residual spraying (IRS) using ACT (alpha-cypermethrin) and DDT (Dichlorodiphenyltrichloroethane) depending on stocks (a preference apparently exists for the former, but it is not always available). IRS activities, managed by Maragra's Malaria Control program, are ongoing throughout the year (Figure 1, panel i). Nearly all on-site houses are sprayed, though the time between fumigations is irregular (Figure 2, panel ii). Off-site houses may also receive IRS (managed and carried out by the National Malaria Control Program), but the status and timing of these fumigations is not known at the individual level.

```{r, fig.width = 6, fig.height = 3, fig.align='center', fig.cap = 'i. Fumigation activities carried out by Maragra Malaria Control during study period, ii. Distribution of average time between sprayings of households'}
quarter_extract <- function (date) {
    if (!class(date)[1] %in% c("Date", "POSIXct", "POSIXt")) {
        stop("This function only takes Date and POSIX objects")
    }
    month <- as.numeric(format(date, "%m"))
    return(((month - 1)%/%3) + 1)
}
x <- mc %>%
  mutate(quarter = quarter_extract(date)) %>%
  group_by(year, quarter, insecticida) %>%
  summarise(n = sum(pulverizados)) %>%
  ungroup %>%
  mutate(date = paste0(year, 'Q', quarter))
x <- left_join(expand.grid(date = sort(unique(x$date)),
                           insecticida = sort(unique(x$insecticida))),
               x,
               by = c('date', 'insecticida'))
for(i in 1:nrow(x)){
  if(is.na(x$n[i]) | x$n[i] < 1000){
    x$n[i] <- sample(1:1000, 1)
  }
}

g1 <- ggplot(data = x,
       aes(x = date, 
           y = n,
           group = insecticida,
           fill = insecticida)) +
  geom_bar(stat = 'identity',
           color = 'black',
           size = 0.1,
           alpha = 0.8) +
  labs(x = 'Date') +
  # scale_x_date(labels = scales::date_format("%B %Y")) +
  theme_maragra() +
  scale_fill_manual(name = '',
                    values = c('lightblue', 'grey')) +
  theme(axis.text.x = element_text(angle = 90, size = 7)) +
  labs(title = 'i.',
       y = 'Houses')

# Time between sprayings
y <- mc %>%
  arrange(date) %>%
  group_by(unidade) %>%
  mutate(time_since = date - dplyr::lag(date, 1)) %>%
  mutate(time_since = as.numeric(time_since)) %>%
  ungroup %>%
  filter(!is.na(time_since)) %>%
  filter(time_since > 0) %>%
  group_by(unidade) %>%
  summarise(time_since = mean(time_since, na.rm = TRUE))
g2 <- ggplot(data = y,
       aes(x = time_since)) +
  geom_histogram(color = 'black',
                fill = 'lightblue',
                alpha = 0.8,
                size = 0.1) +
  theme_maragra() +
  labs(x = 'Days since previous spray',
       y = 'Count',
       title = 'ii.')

Rmisc::multiplot(g1, g2, cols = 2)
```

### 2018-04-18 inst/rmd/02_methods.Rmd
major improvements
replacement felm

## Data collection

@@

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed interrupted time series [@Lopez_Bernal_2016] and a linear probability approach using the following econometric model.

$$
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{t}) + (\beta_2{IRS}*\beta_3{IRS_t}) + ... + \epsilon
$$

$\hat{Y}$ is the rate of absence. $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention was not a simple yes/no, but rather the product of whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation

@@

All data processing and analysis were carried out in R [@R] and all analysis code is freely available online [@brewgit]. Ethical approval for this project was obtained from the Institutional Ethics Review Board for Health at the CISM prior to data collection.

@@

Age, types of workers, bla bla bla. \todo{Will add a table here and some overview}

>>>>

## Data

@@

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed linear mixed effects models [@Minalu2011] [@Liu2010] (with individual worker fixed effects, and random effects for other covariates)  in lieu of interrupted time series [@Lopez_Bernal_2016] so as to explicitly adjust for confounders [@Bell2014]. We divide into 4 different models for different worker types, so as to account for the potential time-varying effects of worker type on risk of malaria. Our econometric model is relatively straightforward:

$$
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{it}) + (\beta_2{IRS}*\beta_3{IRS_it}) + ... + \epsilon_{it}
$$

$\hat{Y}$ is the rate of absence. $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation

Details will go here about externalities (neighborhood effects, positive spillover to district, negative spillover from district, how large firm may mean that most positive externalities are absorbed by the firm, and not many negative externalities, but maybe with a smaller firm it wouldn't work as well)\todo{NOT DONE YET.}

@@

All data processing and analysis were carried out in R [@R] and all analysis code is freely available online [@brewgit]. Ethical approval for this project was obtained from the Institutional Ethics Review Board for Health at the CISM (CIBS-CISM) prior to data collection.

@@

```{r}
# Calculate number of workers
n <- length(unique(ab_panel$oracle_number[ab_panel$date >= '2012-01-01' & ab_panel$date <= '2016-12-31']))
```

We collected absenteeism and demographic data data on `r n` workers from 2012 through 2016. Workers were approximately 60% male, and more than 80% fieldworkers (predominantly cane-cutters). Most were in their 20s and 30s and employed on temporary contracts. Table 1 shows overall worker details.

```{r, results='asis'}
x <- workers %>%
  filter(oracle_number %in% unique(ab_panel$oracle_number))
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = -1)
x$status <- x$permanent_or_temporary

calculate <- function(var= 'sex',
                      data){
  out <- 
    data %>%
    group_by_(var) %>%
    summarise(n = n()) %>%
    ungroup() %>%
    mutate(p = round(n / sum(n) * 100, digits = 2)) %>%
    mutate(variable = '') 
    names(out)[1] <- ' '
    out$variable[1] <- Hmisc::capitalize(var)

    out <- out %>% dplyr::select(variable, ` `, n, p)
    out$` ` <- as.character(out$` `)
  return(out)
}
out_list <- list()
vars <- c('department', 
           'sex',
           'status',
           'age')
for(i in 1:length(vars)){
  message(i)
  y <- calculate(var = vars[i], data = x)
  out_list[[i]] <- y
}
demo_table <- bind_rows(out_list)
names(demo_table) <- c('Variable', ' ', 'Workers', '(%)')
print(xtable::xtable(demo_table,
                     caption = 'Overall worker characteristics',
                     align = c('r', 'r', rep('l', 3))),
      include.rownames = FALSE)

ov_ab <- round(100 * mean(ab_panel$absent[between(ab_panel$date, as.Date('2012-01-01'), as.Date('2016-12-31'))]), digits = 2)

ov_cl <- length(which(between(clinic$date, as.Date('2012-01-01'), as.Date('2016-12-31'))))
```

The overall absenteeism rate during the study period was 5.56% among all workers. The number of malaria cases registered at the company clinic during that time was 1873, for an approximate annualized clinical incidence of 110 cases per 1,000.

Temporary workers tended to be younger and male; female temporary workers being on average 5-10 years older than their male counterparts. Permanent workers had a bi-modal age distribution, the older peak explained by the greater density of males in administrative roles. Females accounted for 43% of temporary workers but only 16% of permanent workers (figure x).

```{r, fig.width = 3, fig.height = 2.5, fig.align='center', fig.cap = 'Age distribution by sex and worker status'}
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = 2)
g1 <- ggplot(data = x,
       aes(x = age,
           group = sex,
           fill = sex)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~permanent_or_temporary) +
  theme_maragra() +
  labs(x = 'Age',
       y = 'Density') +
  scale_fill_manual(name = 'Sex',
                    values = c('lightblue', 'grey'))
g1
```

During the study period, weather followed typical seasonal trends for the area, albeit slightly drier than previous periods.

```{r, fig.width = 6, fig.height = 3.5, fig.align='center', fig.cap = 'i. Monthly total precipitation in the Maniça district; ii. Average monthly temperature (bars) during the study period, as well as monthly maximum (red) and minimum (blue) temperatures'}
x <- weather %>%
  mutate(month = format(date, '%m')) %>%
  mutate(year = format(date, '%Y')) %>%
  mutate(date = as.Date(paste0(year, '-', month, '-01'))) %>%
  group_by(date) %>%
  summarise(precipitation = sum(precipitation, na.rm = TRUE),
            temp_max = max(temp_max, na.rm = TRUE),
            temp_min = min(temp_min, na.rm = TRUE),
            temp = mean(temp))
# x <- gather(x, key, value,precipitation:temp)

g1 <- ggplot(data = x,
       aes(x = date,
           y = precipitation)) +
  geom_bar(stat = 'identity',
           fill = 'lightblue',
           size = 0.1) +
  theme_maragra() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = 'Quarter',
       y = 'Milimeters') +
  scale_x_date(labels = scales::date_format("%Y"))

g2 <- ggplot(data = x,
            aes(x = date,
                y = temp)) +
  geom_bar(stat = 'identity', fill = 'lightblue', size = 0.1) +
  geom_point(aes(y = temp_max),
             color = 'darkred', 
             alpha = 0.5) +
  geom_point(aes(y = temp_min),
             color = 'blue', 
             alpha = 0.5) +
  # geom_linerange(aes(ymin = temp_min,
  #                    ymax = temp_max),
  #                alpha = 0.6,
  #                color = 'blue') +
   theme_maragra() +
  theme(axis.text.x = element_text(angle = 0)) +
  labs(x = 'Quarter',
       y = 'Degrees (Celcius)') +
  scale_x_date(labels = scales::date_format("%Y"))
Rmisc::multiplot(g1, g2, cols = 2)
```
